---
title: "qPCR Data Visualization: BAL Methods Optimization "
output: html_notebook
---
This notebook is for visualizing qPCR data for initial BAL Methods Optmization project. The goal of this notebook is to visualize the following

1. Absolute DNA Concentration per Treatment 
2. Relative (Fold Change to Native) DNA Concentration per Treatment
3. Relative (Fold Change to Native) Human DNA per Nanogram Total DNA
4. Relative (Fold Change to Native) Bacterial DNA per Nanogram Total DNA


First, load packages
```{r}
library(tidyverse)
library(readxl)
library(ggplot2)
library(ggpubr)
library(cowplot)
library(ggbreak) 
library(ggprism)
```


Load a) concentration and b) qPCR data

```{r}
qubit_data <-
  read_excel("~/Documents/northwestern/01_research/bal_methods_optimization/01_data/qpcr/qubit_data.xlsx")

BMOQ_metadata <- read_excel("~/Documents/northwestern/01_research/bal_methods_optimization/02_metadata/BMOQ_metadata.xlsx")

cq_results_18s = 
  read_csv("~/Documents/northwestern/01_research/bal_methods_optimization/01_data/qpcr/BMQ1/02022022/jacksumner_bmoE001-002_2022-02-03 18-43-25_JISQPCR -  Quantification Cq Results_0.csv")

cq_results_16s = 
  read_csv("~/Documents/northwestern/01_research/bal_methods_optimization/01_data/qpcr/BMQ1/02102022/JackSumner_BMOQ1.1_16S_2022-02-11 18-13-29_JISQPCR -  Quantification Cq Results_0.csv")


```
So, from the qPCR data file we are primarily interested in the following columns:

1. Sample 
2. Cq Mean
And we want to use those data and the metadata to calculate a relative change to Native

Therefore, a function to clean and aggregate data is necessary

```{r}
clean_qpcr_data <- function(qpcr_data){
  clean_qpcr_data <- 
    qpcr_data %>% 
    rename(SQ_mean = "SQ Mean") %>% 
    select(Sample, SQ_mean) %>%
    unique() %>% drop_na()
  return(clean_qpcr_data)
}

merge_metadata <- function(qpcr_data_clean, metadata){
  merged_qpcr_data <- 
    metadata %>% 
    full_join(qpcr_data_clean) %>%
    filter(Is_bal == "TRUE") %>%
    select(!Is_bal)
  return(merged_qpcr_data)
}
  
calculate_fold_change <- function(merged_qpcr_data){
  processed_qpcr_data <- merged_qpcr_data %>% 
  filter(Treatment == "Native") %>%
  select(SQ_mean, Bal_id) %>%
  rename(SQ_mean_native = SQ_mean) %>% 
  full_join(merged_qpcr_data, by = "Bal_id") %>%
  mutate(fc = SQ_mean / SQ_mean_native) %>% 
  select(!SQ_mean_native)
  return(processed_qpcr_data)
}

calculate_cq_per_nanogram <- function(processed_qpcr_data, qubit_data){
  merged_df <- left_join(processed_qpcr_data, qubit_data)
  merged_df$qubit_conc <- as.double(merged_df$qubit_conc)
  merged_df <- merged_df %>% mutate(SQ_mean_nanogram = SQ_mean / qubit_conc)
  processed_merged_df <- merged_df %>%
  filter(Treatment == "Native") %>%
  select(SQ_mean_nanogram, Bal_id) %>%
  rename(SQ_mean_native = SQ_mean_nanogram) %>% 
  full_join(merged_df, by = "Bal_id") %>%
  mutate(fc_nanogram = SQ_mean_nanogram / SQ_mean_native) %>% 
  select(!SQ_mean_native)
  return(processed_merged_df)
}

process_data <- function(qpcr_data, metadata, qubit_data){
  qpcr_data_clean <- clean_qpcr_data(qpcr_data)
  merged_qpcr_data <- merge_metadata(qpcr_data_clean, metadata)
  processed_qpcr_data <- calculate_fold_change(merged_qpcr_data)
  processed_qpcr_data_nanogram <- calculate_cq_per_nanogram(processed_qpcr_data, qubit_data)
  return(processed_qpcr_data_nanogram)
}
```

```{r}
processed_18s <- process_data(cq_results_18s, BMOQ_metadata, qubit_data)

processed_18s$Treatment <- 
  factor(processed_18s$Treatment, levels = c("Native","Supernatant", "DNase"))

p1 <- ggplot(processed_18s, aes(x = Treatment, y = fc)) +
  geom_boxplot() + 
  geom_line(aes(group = Bal_id), size = 0.05, alpha = 0.7) +
  ylab("Human DNA (18S gene amplification)") + 
  ggtitle("Fold Change in Human DNA", "Normalized by volume (copies/µL)") + 
  theme_prism(base_fontface = "plain") + 
  ylim(0, 5)
p1
```


Process 16S data 
```{r}
processed_16s <- process_data(cq_results_16s, BMOQ_metadata, qubit_data)

processed_16s$Treatment <- 
  factor(processed_16s$Treatment, levels = c("Native","Supernatant", "DNase"))

p2 <- ggplot(processed_16s, aes(x = Treatment, y = fc)) +
  geom_boxplot() + 
  geom_line(aes(group = Bal_id), size = 0.05, alpha = 0.7) +
  ylab("Bacterial DNA (16S gene amplification)") + 
  ggtitle("Fold Change in Bacterial DNA ", "Normalized by volume (copies/µL)") + 
  theme_prism(base_fontface = "plain") + 
  ylim(0, 5) #+ scale_y_break(c(4, 8), scales = 0.2, ticklabels=c(8.5))
p2

```
Now plot fold change with with data normalized by mass. 

```{r}
# human
p3 <- ggplot(processed_18s, aes(x = Treatment, y = fc_nanogram)) +
  geom_boxplot() + 
  geom_line(aes(group = Bal_id), size = 0.05, alpha = 0.7) +
  ylab("Human DNA (18S gene amplification)") + 
  ggtitle("Fold Change in Human DNA", "Normalized by Mass (copies/ng)") + 
  theme_prism(base_fontface = "plain") + 
  ylim(0, 5)
p3

# bacterial
p4 <- ggplot(processed_16s, aes(x = Treatment, y = fc_nanogram)) +
  geom_boxplot() + 
  geom_line(aes(group = Bal_id), size = 0.05, alpha = 0.7) +
  ylab("Bacterial DNA (16S gene amplification)") + 
  ggtitle("Fold Change in Bacterial DNA", "Normalized by Mass (copies/ng)") + 
  theme_prism(base_fontface = "plain") + 
  ylim(0, 5) #+ scale_y_break(c(4, 18), scales = 0.2, ticklabels=c(18.5))
p4
```


And now, plot all as a panel. Adjust size with ggsave to prevent overlap
```{r}
p5 <- plot_grid(p1, p2, p3, p4) 
p5
```

```{r}
ggsave("../05_figures/01_bmoq001-0002_foldchange_human_volume.pdf", p1)
ggsave("../05_figures/02_bmoq001-0002_foldchange_bacterial_volume.pdf", p2)
ggsave("../05_figures/03_bmoq001-0002_foldchange_human_mass.pdf", p3)
ggsave("../05_figures/04_bmoq001-0002_foldchange_bacterial_mass.pdf", p4)
ggsave("../05_figures/05_bmoq001-0002_foldchange_panel.pdf", p5, height = 10, width =8 , units = "in")

```


```{r}
qubit_data_clean <- qubit_data %>% right_join(BMOQ_metadata) %>% filter(Is_bal == "TRUE")
qubit_data_clean$qubit_conc <- as.double(qubit_data_clean$qubit_conc)
qubit_data_clean$Treatment <- factor(qubit_data_clean$Treatment, level=c("Native", "Supernatant", "DNase"))
p6 <- ggplot(qubit_data_clean, aes(x=Treatment, y=qubit_conc)) +
  geom_boxplot() + 
  ylab("DNA Concentration (ng/µL)") +
  geom_line(aes(group = Bal_id), size = 0.05, alpha = 0.7) +
  ggtitle("DNA Concentration by Treatment") +
  theme_prism(base_fontface = "plain")
p6
ggsave("../05_figures/06_bmoq001-0002_DNA_Concentration.pdf", p6)

```



